---
title: "Vacc vs. Pre pathway enrichment"
author: "Tu Hu"
format: html
editor: visual
---

## Load packages and data

```{r packages}
library(dplyr)
library(tidybulk)
library(fgsea)
library(purrr)
library(ggplot2)
```

## DEG list

```{r}
DEG <- readRDS("../data/DElist/DGE_immu_vs_pre.rds")
str(DEG, max.level = 2)
```

## Gene set

```{r}
gs <- 
  tibble(gs_name = c("CP:REACTOME", 
                     "GO:BP", 
                     "GO:MF")) %>% 
  mutate(gs_list = purrr::map(gs_name, function(subcat){
  gs_df <- msigdbr::msigdbr(species = "Homo sapiens", 
                            subcategory = subcat) %>% 
    group_by(gs_name) %>% tidyr::nest() %>% 
    mutate(gene_id = purrr::map(data, ~ .x %>% pull(gene_symbol))) %>% select(-data)
  gs_list <- gs_df$gene_id
  names(gs_list) <- gs_df$gs_name
  return(gs_list)
}))
```

## pathway enrichment by fgsea

```{r}
design <- 
  tidyr::expand_grid(
    gs_name  = gs$gs_name, 
    vacc = names(DEG),
    contrast = c("immu1_vs_pre", "immu2_vs_pre")
    )

result <- 
  design %>% 
  mutate(
    gs = map(gs_name, function(x){
      gs <- gs[gs$gs_name == x, ]$gs_list[[1]]
      return(gs)
      }),
    DEG = map2(vacc, contrast, 
               ~ DEG[[.x]][[.y]] %>% 
                 as_tibble(rownames = "gene_name") %>% 
                 filter(padj < 0.05) %>% 
                 select(gene_name, log2FoldChange) %>% 
                 tibble::deframe()),
    pathway_enrichment =
      map2(gs, DEG, ~ fgsea(pathways = .x, stats = .y) %>% 
             as_tibble() %>% 
             filter(padj < 0.05) %>% 
             arrange(-NES)) 
    )

saveRDS(result, "../data/pathway_enrichment/GSEA_enrichment.rds")
# result <- readRDS("../data/pathway_enrichment/GSEA_enrichment.rds")

gobp_enrichment <- 
  result %>% filter(gs_name == "GO:BP") %>% pull(pathway_enrichment)

names(gobp_enrichment) <- 
  result %>% filter(gs_name == "GO:BP") %>% 
  mutate(name = paste(vacc, contrast, sep = "_")) %>% 
  pull(name)

openxlsx::write.xlsx(gobp_enrichment, 
                     file = "../data/pathway_enrichment/GSEA_enrichment_gobp.xlsx")
```

-   Reference gene set:
    -   CP:REACTOME (Reactome Pathways): 1615 pathways
    -   GO:BP (Gene Ontology Biological Process): 7658 pathways
    -   GO:MF (Gene Ontology Molecular Function): 1738 pathways
-   Pros and Cons:
    -   GO:BP: comprehensive, but a lot of redundancy.
    -   For example, the top 10 pathways of H107_CAF09c immu1_vs_pre are:
    -   "GOBP_DEFENSE_RESPONSE_TO_SYMBIONT" "GOBP_NEGATIVE_REGULATION_OF_VIRAL_PROCESS"\
        \[3\] "GOBP_TYPE_I\_INTERFERON_PRODUCTION" "GOBP_RESPONSE_TO_VIRUS"\
        \[5\] "GOBP_REGULATION_OF_VIRAL_PROCESS" "GOBP_RESPONSE_TO_TYPE_I\_INTERFERON"\
        \[7\] "GOBP_POSITIVE_REGULATION_OF_TYPE_I\_INTERFERON_PRODUCTION" "GOBP_NEGATIVE_REGULATION_OF_VIRAL_GENOME_REPLICATION"\
        \[9\] "GOBP_INTERFERON_BETA_PRODUCTION" "GOBP_CYTOKINE_MEDIATED_SIGNALING_PATHWAY"
-   No of enriched pathways:
    -   GO:BP
        -   H107_CAF09c
            -   immu1_vs_pre: 313
            -   immu2_vs_pre: 232
        -   H107_CAF04_CpG
            -   immu1_vs_pre: 68
            -   immu2_vs_pre: 95

```{r}
pathway_plot <- 
  readxl::read_excel("../data/gene_set.xlsx") %>% 
  dplyr::rename(pathway = GO_term)
```

```{r Bubble plot}
t_bubble <- 
  result %>% 
  dplyr::select(gs_name:contrast, pathway_enrichment) %>% 
  unnest(cols = "pathway_enrichment") %>% 
  inner_join(pathway_plot) %>% 
  mutate(`-log10 P-value` = -log10(padj))
         # pathway_show = forcats::fct_reorder(pathway_show, NES)) 

g_bubble <- 
  t_bubble %>% 
  ggplot(aes(x = contrast, y = pathway_display, color = NES, size = `-log10 P-value`)) +
  geom_point() +
  facet_grid(. ~ vacc, scales = "free", space = "free") +
  # rotate x xaixs labels
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_size_area(max_size = 8) + 
  scale_color_gradient2(low="blue", mid = "white", high="red") +
  theme_classic() +
  labs(y = NULL)

```

## Bubble plot

```{r}
g_bubble
```
